---
title: "Effects of Timbre on Tonality Perception - Data Analyses"
author: "Eva Wu"
output: 
  github_document:
    toc: true
---

*This script referenced code from my stats professor Liz Page-Gould and collaborator/MA thesis advisor Stephen Van Hedger.*

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(rstatix) # descriptives
library(lme4) # GLMM (GLM w/ random effects)
library(car) # anova test
library(ordinal) # CLMM (ordinal DV)
library(mediation)
library(emmeans) # post hoc
library(performance) # calc ICC 
library(MixedPsy) # PSE
library(ggeffects)
```

This project examines how instrumental timbre influences categorization of three-note arpeggios as major versus minor. Middle note of the arpeggios has five steps, ranging from +0c (minor) to +100c (major) in 25c increments. Participants heard five instruments in total in Experiment 1 (randomized across trial): piano, oboe, violin, trumpet, xylophone. To investigate whether effects were due to acoustic reasons or learned associations, we conducted Experiment 2 using six artificial timbres, varying in terms of spectral centroid (low, mid, high) and amplitude envelope (rounded versus percussive).

Judgments should be theoretically independent of instrument. However, if certain timbres are more strongly associated with positive/negative affect, then it is possible that this will be reflected in the categorization task (i.e., “happier” instruments leading to a greater likelihood of responding “major”), despite no explicit instructions to base judgments on these characteristics.

At the end of the experiment, participants are asked to make explicit ratings on how likely each instrument is to play a sad melody.

```{r import, include = FALSE}
# import cleaned Qualtrics data
df_demo <- read_csv("qualtrics_cleaned.csv") %>%
  mutate(Inst = factor(Inst, levels = c(0, 1), labels = c("No", "Yes")),
         Inst_now = factor(Inst_now, levels = c(0, 1), labels = c("No", "Yes")),
         Read = factor(Read, levels = c(0, 1), labels = c("No", "Yes")),
         Ens = factor(Ens, levels = c(0, 1), labels = c("No", "Yes")),
         Course = factor(Course, levels = c(0, 1), labels = c("No", "Yes")),
         Perf = factor(Perf, levels = c(0, 1, -1), labels = c("No", "Yes", "Not sure")),
         block_passed_practice = factor(block_passed_practice))

# define instrument lists (to avoid redundant code)
instruments1 <- c("oboe", "violin", "trumpet", "piano", "xylophone")
instruments2 <- c("T1", "T2", "T3", "T4", "T5", "T6")

# import tonality categorization data from jspsych
df_cat1 <- read_csv("inst-cat-uc-1.csv") %>%
  filter(designation == "MAIN-JUDGMENT") %>%
  dplyr::select(participant, chord, qualtrics_id, selected_major, instrument, tuning_step) %>%
  mutate(chord = factor(chord),
         instrument = factor(instrument, levels = instruments1),
         # centre so that tuning=3 becomes reference; this is equivalent to group-mean centering bc design is the same across subs
         tuning_c = tuning_step-3) %>%
  semi_join(df_demo, by = c("participant" = "jspsych_id")) # let go of the j sub who didn't have q response

# aggregate data for modeling (mixpsy, the package that helps us calculate pse, can only take count data as DV)
df_sum1 <- df_cat1 %>%
  group_by(participant, chord, tuning_c, instrument) %>%
  summarize(count_major = sum(selected_major),
            count_minor = 8-count_major) %>%
  inner_join(df_demo, by = c("participant" = "jspsych_id")) %>%
  rename(qualtrics_id = participant.y)

# 200 trials per subject (8 repetitions of the same stimulus), 52 subjects total
# treating tuning step as continuous, numeric

# now do the same for exp2
df_cat2 <- read_csv("inst-cat-uc-2.csv") %>%
  filter(designation == "MAIN-JUDGMENT") %>%
  dplyr::select(participant, chord, qualtrics_id, selected_major, instrument, envelope, harmonics, tuning_step) %>%
  mutate(selected_major = as.numeric(selected_major),
         tuning_step = as.numeric(tuning_step),
         chord = factor(chord),
         instrument = factor(instrument, levels = instruments2),
         tuning_c = tuning_step-3)
# 240 trials per subject (8 repetitions of the same stimulu), 58 subjects total

df_sum2 <- df_cat2 %>%
  group_by(participant, chord, tuning_c, instrument, envelope, harmonics) %>%
  summarize(count_major = sum(selected_major),
            count_minor = 8-count_major) %>%
  mutate(envelope = factor(envelope),
         harmonics = factor(harmonics, levels = c("low", "mid", "high"))) %>%
  inner_join(df_demo, by = c("participant" = "jspsych_id")) %>%
  rename(qualtrics_id = participant.y)

# separate exp1&2 qualtrics participants
df_demo1 <- df_demo %>%
  semi_join(df_sum1, by = c("jspsych_id" = "participant"))

df_demo2 <- df_demo %>%
  semi_join(df_sum2, by = c("jspsych_id" = "participant"))

# import explicit valence rating data from jspsych
df_rtg1 <- read_csv("inst-cat-uc-1.csv") %>%
  filter(designation == "INST-VALENCE-RTG") %>%
  dplyr::select(participant, chord, instrument, explicit_rtg)  %>%
  mutate(valence = factor(explicit_rtg),
         instrument = factor(instrument, levels = instruments1)) %>%
  semi_join(df_demo, by = c("participant" = "jspsych_id")) # let go of the j sub who didn't have q response

df_rtg2 <- read_csv("inst-cat-uc-2.csv") %>%
  filter(designation == "INST-VALENCE-RTG") %>%
  dplyr::select(participant, chord, instrument, envelope, harmonics, explicit_rtg)  %>%
  mutate(explicit_rtg = as.numeric(explicit_rtg),
         valence = factor(explicit_rtg),
         instrument = factor(instrument, levels = instruments2),
         envelope = factor(envelope),
         harmonics = factor(harmonics, levels = c("low", "mid", "high"))) %>%
  semi_join(df_demo, by = c("participant" = "jspsych_id"))

# combine cat & rat data
df_combo1 <- left_join(df_sum1, df_rtg1, by = c("participant", "instrument", "chord"))
df_combo2 <- left_join(df_sum2, df_rtg2, by = c("participant", "instrument", "chord", "envelope", "harmonics"))
```

## 1. Demographics

```{r descriptives, warning = FALSE}
# exp1
df_demo1 %>%
  get_summary_stats(c(Age, Inst_yr, accuracy), type = "full") %>%
  dplyr::select(n, variable, mean, sd, min, max)

lapply(df_demo1[c("Gender", "Major", "Inst", "Inst_now", "Read", "Ens", "Course", "Perf", "Time_make", "Time_listen", "Concert", "block_passed_practice")], table)

df_demo1 %>%
  ggplot(aes(Age)) +
  geom_bar()

df_demo1 %>%
  ggplot(aes(Inst_yr)) +
  geom_bar() +
  labs(title = "Years of most trained instrument")

# exp2
df_demo2 %>%
  get_summary_stats(c(Age, Inst_yr, accuracy), type = "full") %>%
  dplyr::select(n, variable, mean, sd, min, max)

lapply(df_demo2[c("Gender", "Major", "Inst", "Inst_now", "Read", "Ens", "Course", "Perf", "Time_make", "Time_listen", "Concert", "block_passed_practice")], table)

df_demo2 %>%
  ggplot(aes(Age)) +
  geom_bar()

df_demo2 %>%
  ggplot(aes(Inst_yr)) +
  geom_bar() + 
  labs(title = "Years of most trained instrument")
```

48 participants in exp1, 58 in exp2.

Most of our sample had prior musical training. This is good because together with the practice test results, we were quite confident that people knew what they were supposed to do. However, this limits the generalizability of our results to non-musicians.

## 2. Tonality Categorization Analysis

We used binomial/probit regression to model tonality categorization against instrument timbre and tuning step, because DV is binary (major = 1, minor = 0), and we were interested in comparing points of subjective equality (PSEs) for each instrument (i.e., tuning step at which participants' likelihood of choosing major/minor is the same).

### 2.1 Calculate ICC

```{r cat-icc}
# baseline model
model_base_cat1 <- df_sum1 %>%
  glmer(cbind(count_major, count_minor) ~ (1 | participant) + 1, family = binomial(link = "probit"), data=.)
icc(model_base_cat1) # from performance package

model_base_cat2 <- df_sum2 %>%
  glmer(cbind(count_major, count_minor) ~ (1 | participant) + 1, family = binomial(link = "probit"), data=.)
icc(model_base_cat2)
```

First, we checked for ICC of the baseline model. ICC for both experiments were greater than .1, indicating clustering effects within each participant. Therefore, random effects were included in the models.

### 2.2 Model Comparison (GLMM)

We modeled probability of major tonality categorization against tuning step (5 levels, treated as a continuous numeric variable, centered ahead of time) and instrument timbre (5 levels for exp1, 6 for exp2, categorical variable). 

We were not sure which random slopes to include, if at all, so we compared models without random slope (only random intercept), with random slope of instrument or tuning step, and with both random slopes.

```{r cat-model-comparison, include = FALSE}
# start with maximal model (Barr et al., 2013) 
# since we are using mixpsy to calculate PSE, we have to convert binary DV to sum
model_maximal1 <- df_combo1 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * instrument + (1 + tuning_c + instrument | participant), 
        family = binomial(link = "probit"), data = ., control = glmerControl(optimizer = "bobyqa"))

# model is singular (convergence issue) - model too complicated
# which makes sense given instrument is a categorical variable
model_maximal2 <- df_combo2 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * instrument + (1 + tuning_c + instrument | participant), 
        family = binomial(link = "probit"), data = ., control = glmerControl(optimizer = "bobyqa"))

# try deleting 1 random slope
model_tuning1 <- df_combo1 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * instrument + (1 + tuning_c | participant), 
        family = binomial(link = "probit"), data = ., control = glmerControl(optimizer = "bobyqa"))

model_tuning2 <- df_combo2 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * instrument + (1 + tuning_c | participant), 
        family = binomial(link = "probit"), data = ., control = glmerControl(optimizer = "bobyqa"))

# try deleting another random slope instead
model_instrument1 <- df_combo1 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * instrument + (1 + instrument | participant), 
        family = binomial(link = "probit"), data = ., control = glmerControl(optimizer = "bobyqa"))

# model is still singular (too many instruments)
model_instrument2 <- df_combo2 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * instrument + (1 + instrument | participant), 
        family = binomial(link = "probit"), data = ., control = glmerControl(optimizer = "bobyqa"))

anova(model_maximal1, model_tuning1)
# including both random slope (vs only 1 of the 2) significantly improves model
anova(model_maximal1, model_instrument1)
# including random slope of tuning step (vs instrument) results in lower AIC/BIC 
anova(model_tuning1, model_instrument1)
```

Even though likelihood ratio test (LRT) showed that the model including both random slopes explained the most variance, it resulted in model convergence issues (especially for exp2 because there were 6 instruments). Therefore, we decided to only include 1 random slope. Including the random slope for tuning step resulted in a better model than including that for instrument. 

### 2.3 Model Summary

```{r cat-model-summary}
# examine results of the model we decided on
summary(model_tuning1) # baseline = oboe
summary(model_tuning2) # baseline = T1

Anova(model_tuning1, type = 3) # sig main effects, non-sig interaction
Anova(model_tuning2, type = 3) # sig main effects & interaction
```
|                             | Exp1 | Exp2 |
| --------------------------- | ---- | ---- |
| Main effect of timbre       | ✓    | ✓    |
| Main effect of tuning       | ✓    | ✓    |
| Timbre × Tuning interaction | ✗    | ✓    |

Sig main effects for tuning & instrument for both experiments, but interaction was non-sig for exp1 but sig for exp2. In exp1, the effect was primarily expressed as a shift in the categorization boundary (PSE), while in exp2, timbre altered listeners' sensitivity to tuning manipulations.

### 2.4 Post-hoc Comparison

```{r cat-posthoc}
# pairwise comparison at the 3rd tuning_step (+50c)
# exp1
emmeans(model_tuning1, "instrument") %>%
  pairs()

# exp2
emmeans(model_tuning2, "instrument") %>%
  pairs()
```

At the 3rd tuning step (+50c) - 

Exp1: oboe < all, xylophone > all

Exp2: T6 > all but T3, T3 > T1&2, T1 < T3&6

### 2.5 PSE Analysis

PSE represents the tuning step at which the probability of selecting major (or minor) is .5 (probit(p) = 0). Higher PSE means chords played on that instrument tend to sound more minor. Lower PSE means the instrument is more biased towards major.

```{r pse}
# the mixpsy package can helps us achieve this with 1 line of code
mixpsy_x1 <- xplode(model_tuning1, name.cont="tuning_c", name.factor = "instrument")
pses1 <- MixDelta(mixpsy_x1) 

mixpsy_x2 <- xplode(model_tuning2, name.cont="tuning_c", name.factor = "instrument")
pses2 <- MixDelta(mixpsy_x2) # convergence issue, but gradient only .007 -> acceptable

pses1
pses2
```

Exp1: oboe > all, xylophone < all but piano, middle 3 don't differ

Exp2: only T2 > T6

### 2.6 Visualizations

We calculated predicted probability values at tuning steps with smaller bins so that we could plot a smoothed curve graph rather than just a line graph. We also calculated predicted probability values at the 3rd (middle) tuning step.

```{r df2plot, include = FALSE}
df_cat_plot1 <- emmeans(model_tuning1, c("tuning_c", "instrument"),
                        at = list(instrument = instruments1,
                                  tuning_c = seq(-2, 2, length = 25)), 
                        type = "response") %>% # converts to probability rather than probit
  as_tibble() 

# predicted probability at 3rd tuning step
df_cat_pts1 <- emmeans(model_tuning1, c("tuning_c", "instrument"),
                       at = list(instrument = instruments1,
                                 # discrete tuning steps
                                 tuning_c = -2:2), 
                       type = "response") %>%
  as_tibble()

# do the same for exp2
df_cat_plot2 <- emmeans(model_tuning2, c("tuning_c", "instrument"),
                        at = list(instrument = instruments2,
                                  tuning_c = seq(-2, 2, length = 30)), 
                        type = "response") %>%
  as_tibble() 

df_cat_pts2 <- emmeans(model_tuning2, c("tuning_c", "instrument"),
                       at = list(instrument = instruments2,
                                 # discrete tuning steps
                                 tuning_c = -2:2), 
                       type = "response") %>%
  as_tibble()

# PSEs for plotting
df_pse1 <- pses1 %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  pivot_longer(-parameter,
               names_to = "instrument",
               values_to = "value") %>%
  separate(instrument, c("instrument", "param")) %>%
  filter(parameter == "pse") %>%
  pivot_wider(names_from = "param", values_from = "value") %>%
  mutate(instrument = str_remove(instrument, "instrument"),
         instrument = factor(instrument, levels = instruments1), # reorder
         PSE = (Estimate+2)*25,
         SE = Std*25,
         CI.low = (Inferior+2)*25,
         CI.upp = (Superior+2)*25) %>%
  dplyr::select(-parameter, -Estimate, -Std, -Inferior, -Superior)

df_pse2 <- pses2 %>%
  as.data.frame() %>%
  rownames_to_column("parameter") %>%
  pivot_longer(-parameter,
               names_to = "instrument",
               values_to = "value") %>%
  separate(instrument, c("instrument", "param")) %>%
  filter(parameter == "pse") %>%
  pivot_wider(names_from = "param", values_from = "value") %>%
  mutate(instrument = str_remove(instrument, "instrument"),
         PSE = (Estimate+2)*25,
         SE = Std*25,
         CI.low = (Inferior+2)*25,
         CI.upp = (Superior+2)*25) %>%
  dplyr::select(-parameter, -Estimate, -Std, -Inferior, -Superior)
```

```{r cat-plot}
# exp1
# graphing probability
df_cat_plot1 %>%
  ggplot(aes(x = tuning_c, y = prob, color = instrument)) +
  geom_line() + # so that it reflects model predicted value
  geom_line(aes(y = 0.5), color = "grey", linetype = "dotted") +
  geom_point(data = df_cat_pts1) +
  geom_errorbar(data = df_cat_pts1, aes(ymin = prob-SE, ymax = prob+SE), width = 0.1) +
  scale_x_continuous(labels = c("0", "25", "50", "75", "100")) +
  labs(y = "Predicted likelihood of major categorization",
       x = "Tuning step (+cents)", caption = "error bar = SE") +
  theme_minimal()

# tuning step = 3
df_cat_pts1 %>%
  filter(tuning_c == 0) %>%
  ggplot(aes(instrument, prob, fill = instrument)) +
  geom_col() +
  geom_errorbar(aes(ymin = prob-SE, ymax = prob+SE), width = 0.1) +
  labs(x = "Instrument", y = "Avg predicted likelihood of major categorization",
       title = "Tonality Categorization at Tuning Step = +50c", caption = "error bar = SE") +
  guides(fill = "none") +
  theme_minimal()

# pse
df_pse1 %>%
  ggplot(aes(instrument, PSE, fill = instrument)) +
  geom_col() +
  geom_errorbar(aes(ymin = CI.low, ymax = CI.upp), width = 0.1) + 
  labs(x = "Instrument", y = "Tuning step (cents) at which \nprobability of major categorization = 50%",
       title = "Points of Subjective Equality (PSE)", caption = "error bar = 95% CI") +
  guides(fill = "none") +
  theme_minimal()

# exp2
df_cat_plot2 %>%
  ggplot(aes(x = tuning_c, y = prob, color = instrument)) +
  geom_line() + # so that it reflects model predicted value
  geom_line(aes(y = 0.5), color = "grey", linetype = "dotted") +
  geom_point(data = df_cat_pts2) +
  geom_errorbar(data = df_cat_pts2, aes(ymin = prob-SE, ymax = prob+SE), width = 0.1) +
  scale_x_continuous(labels = c("0", "25", "50", "75", "100")) +
  labs(y = "Predicted Likelihood of Major Categorization",
       x = "Tuning step (+cents)", caption = "error bar = SE") +
  theme_minimal()

df_cat_pts2 %>%
  filter(tuning_c == 0) %>%
  ggplot(aes(instrument, prob, fill = instrument)) +
  geom_col() +
  geom_errorbar(aes(ymin = prob-SE, ymax = prob+SE), width = 0.1) +
  labs(x = "Instrument", y = "Avg predicted likelihood of major categorization",
       title = "Tonality Categorization at Tuning Step = +50c", caption = "error bar = SE") +
  guides(fill = "none") +
  theme_minimal()

df_pse2 %>%
  ggplot(aes(instrument, PSE, fill = instrument)) +
  geom_col() +
  geom_errorbar(aes(ymin = CI.low, ymax = CI.upp), width = 0.1) + 
  labs(x = "Instrument", y = "Tuning step (cents) at which \nprobability of major categorization = 50%",
       title = "Points of Subjective Equality (PSE)", caption = "error bar = 95% CI") +
  guides(fill = "none") +
  theme_minimal()
```

## 3. Explicit Valence Ratings Analysis

We modeled explicit valence rating for each instrument against instrument timbre. We used Cumulative Link Mixed Models (CLMM) because DV is ordinal. 

### 3.1 Calculate ICC

```{r exp-icc}
model_base_rtg1 <- clmm(valence ~ 1 + (1 | participant), data = df_rtg1)
model_base_rtg2 <- clmm(valence ~ 1 + (1 | participant), data = df_rtg2)
icc(model_base_rtg1)
icc(model_base_rtg2)
```

ICC from both experiments were smaller than .1, but we still chose to employ the nested model due to clustering within individuals.

### 3.2 CLMM

```{r clmm}
# fit clmm b/c rating is ordinal not interval (1-4 likert scale)
model_rtg1 <- clmm(valence ~ instrument + (1 | participant), data = df_rtg1, Hess = TRUE)
summary(model_rtg1)
#"cond. H" shows identifiability of the model. The smaller the better; not ok above 10^4. We are ok.
joint_tests(model_rtg1)
# cumulative odds ratio (OR) relative to baseline (< 1 = less likely, > 1 = more likely)
exp(coef(model_rtg1)[4:7]) # x times more likely to have a higher rating than oboe

# exp2
model_rtg2 <- clmm(valence ~ instrument + (1 | participant), data = df_rtg2, Hess = TRUE)
summary(model_rtg2)
joint_tests(model_rtg2)
exp(coef(model_rtg2)[4:8]) # x times more likely to have a higher rating than T1
```

In both experiments, timbre and valence were related.

### 3.3 Post-hoc Comparison

```{r clmm-posthoc}
# correction: Tukey (default)
emmeans(model_rtg1, ~ instrument) %>%
  pairs()
emmeans(model_rtg2, "instrument") %>%
  pairs()
```

Exp1: oboe < piano, xylophone; violin < trumpet, piano, xylophone

Exp2: T1 < all but T2, T2 < T4-6, T6 > T1-3

### 3.4 Visualizations 

```{r exp-plot}
# exp1
# plot mean rating as estimated from the model
as.data.frame(emmeans(model_rtg1, ~ valence | instrument, mode = "prob")) %>%
  group_by(instrument) %>%
  summarise(mean_rating = sum(as.numeric(valence) * prob)) %>%
  ggplot(aes(reorder(instrument, mean_rating), mean_rating, fill = instrument)) +
  geom_col(position = "dodge") +
  labs(x = "Instrument", y = "Estimated mean valence rating",
       title = "Explicit Emotional Valence Rating for each Instrument") +
  guides(fill = "none") +
  theme_minimal()

# problem: cannot plot aggregated SE
# alternative: use raw data
df_rtg1 %>%
  group_by(instrument) %>%
  summarise(mean_rating = mean(explicit_rtg),
            se_rating = sd(explicit_rtg/sqrt(n()))) %>%
  ggplot(aes(instrument, mean_rating, fill = instrument)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = mean_rating-se_rating, ymax = mean_rating+se_rating), width = .1) +
  labs(x = "Instrument", y = "Mean valence rating", caption = "error bar = SE",
       title = "Emotional Valence Rating for each Instrument") +
  guides(fill = "none") +
  theme_minimal()

# plot estimated probability of each rating (not very intuitive so prolly won't use)
as.data.frame(emmeans(model_rtg1, ~ valence | instrument, mode = "prob")) %>%
  ggplot(aes(valence, prob, group = instrument, color = instrument)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = prob-SE, ymax = prob+SE), width = .1) +
  labs(x = "Explicit valence rating", y = "Probability for each rating level",
       caption = "error bar = SE") +
  theme_minimal()

# exp2
# plot mean rating 
as.data.frame(emmeans(model_rtg2, ~ valence | instrument, mode = "prob")) %>%
  group_by(instrument) %>%
  summarise(mean_rating = sum(as.numeric(valence) * prob)) %>%
  ggplot(aes(reorder(instrument, mean_rating), mean_rating, fill = instrument)) +
  geom_col(position = "dodge") +
  labs(x = "Instrument", y = "Mean valence rating",
       title = "Explicit Emotional Valence Rating for each Instrument") +
  guides(fill = "none") +
  theme_minimal()

# using raw data
df_rtg2 %>%
  group_by(instrument) %>%
  summarise(mean_rating = mean(explicit_rtg),
            se_rating = sd(explicit_rtg/sqrt(n()))) %>%
  ggplot(aes(instrument, mean_rating, fill = instrument)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = mean_rating-se_rating, ymax = mean_rating+se_rating), width = .1) +
  labs(x = "Instrument", y = "Mean valence rating", caption = "error bar = SE",
       title = "Emotional Valence Rating for each Instrument") +
  guides(fill = "none") +
  theme_minimal()

# plot estimated probability
as.data.frame(emmeans(model_rtg2, ~ valence | instrument, mode = "prob")) %>%
  ggplot(aes(valence, prob, group = instrument, color = instrument)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = prob-SE, ymax = prob+SE), width = .1) +
  labs(x = "Explicit valence rating", y = "Probability for each rating level",
        caption = "error bar = SE") +
  theme_minimal()
```

## 4. The Role of Valence

### 4.1 Valence as an Exploratory Variable

We added explicit valence rating as another predictor in our GLMM, to see if this accounts for the timbre-tonality relationship.

```{r model_add_valence}
model_add_valence1 <- df_combo1 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * instrument + explicit_rtg + 
          (1 + tuning_c | participant), family = binomial(link = "probit"), 
        data = ., control = glmerControl(optimizer = "bobyqa"))
model_add_valence2 <- df_combo2 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * instrument + explicit_rtg + 
          (1 + tuning_c | participant), family = binomial(link = "probit"), 
        data = ., control = glmerControl(optimizer = "bobyqa"))

summary(model_add_valence1)
Anova(model_add_valence1, type = 3)
summary(model_add_valence2)
Anova(model_add_valence2, type = 3)

# exp1:
# instrument - tonality sig (c=total effect)
# instrument - valence sig (a)
# instrument - tonality after controlling for valence still sig (c'=direct effect)
# valence - tonality sig (after controlling for timbre) (b=mediation effect)
# potential partial mediation

# exp2:
# instrument - tonality sig (c=total effect)
# instrument - valence sig (a)
# instrument - tonality after controlling for valence still sig (c'=direct effect)
# valence - tonality nonsig (after controlling for timbre) (b=mediation effect)
# no mediation
```

### 4.2 Mediation Analysis

```{r mediation}
set.seed(2026) # for reproducibility
# aggregate prop_major across tuning steps (so that both models use df w/ same nrows)
df_avg1 <- df_combo1 %>%
  group_by(participant, instrument, explicit_rtg) %>%
  summarise(sum_maj = sum(count_major),
            sum_min = sum(count_minor),
            .groups = "drop")

model.m1 <- df_avg1 %>%
  lmer(formula = explicit_rtg ~ instrument + (1 | participant), data = .)
model.y1 <- df_avg1 %>%
  glmer(formula = cbind(sum_maj, sum_min) ~ instrument + explicit_rtg + (1 | participant), family = binomial(link = "probit"), data = .)

med1 <- mediate(model.m1, model.y1, treat = "instrument", mediator = "explicit_rtg", control.value = "oboe", treat.value = "xylophone") 

df_avg2 <- df_combo2 %>%
  group_by(participant, instrument, explicit_rtg) %>%
  summarise(sum_maj = sum(count_major),
            sum_min = sum(count_minor),
            .groups = "drop")

summary(med1)

model.m2 <- df_avg2 %>%
  lmer(formula = explicit_rtg ~ instrument + (1 | participant), data = .)
model.y2 <- df_avg2 %>%
  glmer(formula = cbind(sum_maj, sum_min) ~ instrument + explicit_rtg + (1 | participant), family = binomial(link = "probit"), data = .)

med2 <- mediate(model.m2, model.y2, treat = "instrument", mediator = "explicit_rtg", control.value = "T2", treat.value = "T6") 

summary(med2)
```

| After controlling for valence (GLMM) | Experiment 1 | Experiment 2 |
| ------------------------------ |--------------|--------------|
| Tuning main effect on Tonality | ✓            | ✓            |
| Timbre main effect on Tonality | ✓            | ✓            |
| Timbre × Tuning interaction    | ✗            | ✓            |
| Valence main effect on Tonality | ✓            | ✗            |
| ----------------------------- |--------------|--------------|
| Timbre-Valence relationship (from CLMM) | ✓            | ✓            |

In both experiments, timbre was significantly associated with both tonality and valence. After controlling for valence, timbre still significantly predicted tonality perception, although the magnitude of the timbre effect was attenuated. However, in exp1, valence was also a significant predictor of tonality, after controlling for timbre-related effects, whereas in exp2, that was no longer the case - after controlling for timbre, valence did not significantly predict tonality perception. This pattern suggests that valence may explain part of the timbre-tonality relationship observed in real instruments, but in artificial timbres, this relationship cannot be explained by emotional valence alone.

Experiment 2 partially replicated Experiment 1 results, suggesting that exp1 effects were partly acoustically grounded and partly due to learned associations.

## 5. Separate Timbral Features

In Experiment 2, we created artificial timbres, systematically varying their envelopes and spectral centroids. T1-3 had a rounded envelope, while T4-6 had a percussive envelope. T1&4 had a lower spectral centroid, T2&5 had a mid spectral centroid, while T3&6 had a higher spectral centoid. Here we separated the "intrument" variable into "envelope" and "harmonics" to look at what was driving the effect of timbre. 

```{r model_env*harm}
# Tonality Categorization
model_sep2 <- df_sum2 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * envelope * harmonics + 
          (1 + tuning_c | participant), family = binomial(link = "probit"), 
        data = ., control = glmerControl(optimizer = "bobyqa"))
summary(model_sep2)
emmeans(model_sep2, "envelope") %>%
  pairs()
emmeans(model_sep2, "harmonics") %>%
  pairs()
emmeans(model_sep2, ~ envelope * harmonics) %>%
  pairs()
anova(model_tuning2, model_sep2) # this model did not perform better than our OG GLMM
Anova(model_sep2, type = 3)
plot(emmeans(model_sep2, ~ harmonics * envelope))

# timbre-valence model
model_sep_rtg2 <- clmm(valence ~ envelope * harmonics + (1 | participant), data = df_rtg2)
summary(model_sep_rtg2)
emmeans(model_sep_rtg2, "harmonics") %>%
  pairs()
# main effect: high > low strong; mid > low but p =.044
emmeans(model_sep_rtg2, ~ envelope * harmonics) %>%
  pairs()
joint_tests(model_sep_rtg2) # cuz Anova can't be used for clmm
plot(emmeans(model_sep_rtg2, ~ harmonics * envelope))

# plot 3-way interaction
pred <- ggpredict(
  model_sep2,
  terms = c("tuning_c [-2:2 by=.1]",
            "envelope",
            "harmonics")
)

ggplot(pred,
       aes(x = x,
           y = predicted,
           color = group)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = conf.low,
                  ymax = conf.high,
                  fill = group),
              alpha = .15,
              colour = NA) +
  facet_wrap(~facet) +
  labs(
    x = "Tuning",
    y = "Predicted probability of 'Major'",
    color = "Envelope",
    fill = "Envelope"
  ) +
  theme_classic()

ggplot(pred,
       aes(x = x,
           y = predicted,
           color = facet)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = conf.low,
                  ymax = conf.high,
                  fill = facet),
              alpha = .15,
              colour = NA) +
  facet_wrap(~group) +
  labs(
    x = "Tuning",
    y = "Predicted probability of 'Major'",
    color = "Harmonics",
    fill = "Harmonics"
  ) +
  theme_classic()
```

| Experiment 2                  | Tonality Categorization | Valence Rating |
| ----------------------------- | ----------------------- | -------------- |
| Envelope main effect          | ✗                       | ✓              |
| Harmonics main effect         | ✓                       | ✓              |
| Envelope × Harmonics interaction | ✗                    | ✗              |
| Tuning × Envelope interaction | ✓                       | NA             |
| Tuning × Harmonics interaction | ✗                      | NA             |
| Tuning × Envelope × Harmonics interaction | ✓           | NA             |
| Harmonics post-hoc low-mid difference | ✗               | ~✓ (p = .044)  |
| Harmonics post-hoc mid-high difference | ✓              | ✗              |

Other post-hoc differences: 

- **Tonality Categorization:** rounded low/mid < rounded/percussive high; percussive low/mid < high

- **Explicit Valence Rating:** rounded low < percussive mid; rounded mid < percussive low; rounded low/mid < percussive high 

Similar trend between tonality categorization and explicit valence rating.

## 6. Exploratory Analyses

We added key and music training variables as exploratory predictors in the model. None of them changed the effect of our main IVs of interest, so we left them out of our models for parsimony. These analyses will not be included in main results, but will potentially be included as supplementary materials.

### 6.1 Key

```{r key}
model_key1 <- df_sum1 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * instrument + chord + (1 + tuning_c | participant), 
        family = binomial(link = "probit"), data = ., control = glmerControl(optimizer = "bobyqa"))
anova(model_key1, model_tuning1) # slightly improves model fit
summary(model_key1) 

# now do the same for exp2
# include all possible interactions for key
model_key2 <- df_sum2 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * instrument + chord + (1 + tuning_c | participant), 
        family = binomial(link = "probit"), data = ., control = glmerControl(optimizer = "bobyqa"))
anova(model_key2, model_tuning2) # slightly improves model fit
summary(model_key2) 

# add key to the sep model
model_sep_key2 <- df_sum2 %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * envelope * harmonics * chord + (1 + tuning_c | participant), 
        family = binomial(link = "probit"), data = ., control = glmerControl(optimizer = "bobyqa"))
summary(model_sep_key2)
anova(model_sep_key2, model_sep2) # does not significantly improve model fit

Anova(model_key1, type = 3)
Anova(model_key2, type = 3)
```

### 6.2 Musical Training

We wanted to add Inst_yr, etc., but there were too many NA values, so we couldn't compare it against the OG model.

`Inst/Inst_now` don't have NAs, but they are not balanced (only <10 answered no to Inst / yes to Inst_now) - not enough data to reach stable model params.

### 6.3 Pianists only

We filtered out participants who have had experience playing an instrument that was used as stimuli in exp1.

```{r pianists}
# how many ppl in exp1 have played the instruments used in exp1
# character(0) means played instruments not used in exp1
df_sum1 %>%
  ungroup() %>%
  dplyr::select(participant, Inst_best) %>%
  unique() %>%
  group_by(Inst_best) %>%
  summarise(n())
# 25 piano, 9 violin
# only worth trying it w piano

model_pianists <- df_sum1 %>%
  filter(grepl("piano", Inst_best)) %>%
  glmer(cbind(count_major, count_minor) ~ tuning_c * instrument + (1 + tuning_c | participant), 
        family = binomial(link = "probit"), data = ., control = glmerControl(optimizer = "bobyqa"))

summary(model_pianists)
```

No difference in slopes for piano vs other instruments among pianists.