---
title: "Data Cleaning & Exploratory Graphs"
author: "Eva Wu"
output: github_document
date: "2026-05-12"
---

```{r setup, include = FALSE}
library(tidyverse)
library(colorspace)
library(kableExtra)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Data Cleaning

### Import data

This code chunk is to import concatenated data, delete irrelevant variables and metadata to prepare for further data cleaning. We have data from both jspsych and Qualtrics, and the two were linked by participants' `qualtrics_id` (10-digit random number assigned by Qualtrics; this was copied automatically to jspsych) and their `jspsych_id` (14-digit random letters and numbers assigned at the end of jspsych task; they were asked to manually copy and paste this code to Qualtrics to show that they finished the jspsych task; all but one pasted the right code).

```{r import-qualtrics}
# load
df_q <- read_csv("qualtrics_raw.csv") %>%
  filter(row_number() > 7) %>% # delete pilot data; arranged by RecordedDate by default; recorded before 2/17 = pilot
  select(-c(2:6, 8:16, 50:51)) %>% # delete unnecessary meta data
  mutate(Age = as.numeric(Age),
         Gender = factor(Gender, levels = c(1:3), labels = c("Male", "Female", "Non-binary")),
         Year = factor(Year, levels = c(1:6), labels = c("Freshman", "Sophomore", "Junior", "Senior", "Graduate", "Other")),
         Major = factor(Major, levels = c(1:3), labels = c("Psychology", "Music", "Other")),
         # extract first number as starting age
         Start = as.numeric(sub("^.*?(\\d+).*", "\\1", Start)),
         # manually inspect and correct the wrong one
         Start = if_else(jspsych_id == "tkp19u9btbvhee", 9, Start),
         Pitch = as.integer(`Pitch&Tempo_1`),
         Tempo = as.integer(`Pitch&Tempo_2`),
         # extract first number as number of concerts attended per year
         Concert = as.integer(sub("^.*?(\\d+).*", "\\1", Concert))) %>%
  # calculate the number of years participants played their most-trained instrument  
  rowwise() %>%
  mutate(Inst_yr = max(as.integer(unlist(str_extract_all(Inst_list, "\\d+"))))) %>%
  ungroup() %>%
  select(-`Pitch&Tempo_1`, -`Pitch&Tempo_2`)

# alternative code:
# for (i in 1:length(df_q$Inst_list)) {df_q$Inst_yr[i] <- max(as.integer(unlist(str_extract_all(df_q$Inst_list[i], "\\d+"))))}

# we got lots of -Inf values due to people not listing their years of training
# for those who indicated they are still playing (Inst_now == 1) and filled out a starting age and their current age, I manually calculated it
# for those who did play something but were not playing anymore, I input NA as their score 
# for those who nvr played, I put 0
df_q %>%
  filter(Inst_yr == -Inf)

df_q <- df_q %>%
  # manually correct the wrongly coded ones 
  mutate(Inst_yr = case_when(jspsych_id == "evx5gb3ga2u3qk" ~ 15,
                             jspsych_id == "h21fw4bmslr0vb" ~ 3,
                             jspsych_id == "14vd1s548b9ux8" ~ 5,
                             jspsych_id == "8glm4kxovqneap" ~ 4,
                             jspsych_id == "46ak9p0c90r96o" ~ 1,
                             jspsych_id == "lws6fybx47tsfx" ~ 6,
                             jspsych_id == "cny0e1gcjouwbl" ~ 2,
                             jspsych_id == "r51mv84rhfl9sj" ~ 0.5,
                             jspsych_id == "5qjd88k61078vu" ~ 9,
                             TRUE ~ Inst_yr),
         Inst_yr = ifelse(is.na(Inst_yr), 0, Inst_yr),
         Inst_yr = ifelse(Inst_yr == -Inf, NA, Inst_yr)) %>%
  # flag it if anyone has had training on the instruments used in experiment 1
  mutate(Inst_list = tolower(Inst_list),
         Inst_best = str_extract_all(Inst_list, regex("oboe|piano|violin|trumpet|xylophone")),
         Inst_best = as.character(Inst_best)) %>% # bc can't save list vars in csv
  relocate(Inst_yr, .after = Inst_list) %>%
  relocate(Inst_best, .after = Inst_yr)
```

```{r import-jspsych}
# load exp1 data
df_j1 <- read_csv("inst-cat-uc-1.csv") %>% 
  filter(designation != "NA" & 
         designation != "practice-intro" & 
         designation != "practice-resp" & 
         !qualtrics_id %in% c(NA, 9643222579)) %>% # delete 2 pilot data
  select(participant, qualtrics_id, chord, designation, response, # delete meta data
         correct, passed_practice, block_passed_practice, practice_score, 
         instrument, valence, tuning_step, selected_major, explicit_rtg) %>%
  mutate(qualtrics_id = if_else(participant == "ch4dg75c7th9g1", "1706631024", qualtrics_id),
         instrument = factor(instrument, levels = c("xylophone", "piano", "violin", "trumpet", "oboe")),
         correct = as.numeric(correct),
         practice_score = as.numeric(practice_score),
         tuning_step = as.numeric(tuning_step),
         selected_major = as.numeric(selected_major),
         explicit_rtg = factor(explicit_rtg))

df_j2 <- read_csv("inst-cat-uc-2.csv") %>% 
  filter(designation != "NA" & 
         designation != "practice-intro" & 
         designation != "practice-resp" & 
         participant != "participant") %>% 
  select(participant, qualtrics_id, chord, designation, response, # delete meta data
         correct, passed_practice, block_passed_practice, practice_score, 
         instrument, envelope, harmonics, tuning_step, selected_major, explicit_rtg) %>%
  mutate(instrument = factor(instrument, levels = c("T1", "T2", "T3", "T4", "T5", "T6")),
         correct = as.numeric(correct),
         practice_score = as.numeric(practice_score),
         tuning_step = as.numeric(tuning_step),
         selected_major = as.numeric(selected_major),
         explicit_rtg = as.numeric(explicit_rtg))

# examine j IDs
df_j1_id <- df_j1 %>% 
  select(participant, qualtrics_id) %>% 
  unique() %>% # 49 unique jspsych_id in df_j1
  # for some reason 2 participants' q id on j was "participant", so I recoded one according to their q id on q
  # one of them has 2 q ids (duplicated rows) on q so just leave it as "participant"
  mutate(qualtrics_id = if_else(participant == "ch4dg75c7th9g1", "1706631024", qualtrics_id))

df_j2_id <- df_j2 %>% 
  filter(participant != "participant") %>%
  select(participant, qualtrics_id) %>% 
  unique()
# 58 unique jspsych_id in df_j2; no one's q id was recorded
```

### Deduplicate

Check for duplicated Qualtrics entries and delete one of the duplicates. This might be due to participants closing the Qualtrics tab while completing the jspsych task, and then being redirected to a new Qualtrics page after jspsych task was over.

Decision rule: 
1) if one row has a qualtrics_id recorded on jspsych, then keep that row (only possible in exp1);
2) if one row is blank/partially filled while the other is fully filled, then keep the more filled row;
3) if no qualtrics_id was recorded on jspsych and both rows are filled to the same extent, keep the one with earlier StartDate (this is arbitrarily decided).

```{r deduplicate}
# identify duplicated participants on q by manually examining the data
df_q %>%
  filter(jspsych_id %in% df_j1_id$participant) %>%
  arrange(jspsych_id) # seems like these duplicated rows were both filled and had similar values

# check the 2 duplicated participants' q id on j to determine which row to keep
df_j1_id %>%
  filter(participant == "w3uh3gtrykkjgc" | participant == "ept8xz3drgq38w")

# now do the same for exp2
df_q %>%
  filter(jspsych_id %in% df_j2_id$participant) %>%
  arrange(jspsych_id)

df_j2_id %>%
  filter(participant == "a48zc72k379mm5" | participant == "hecbjunkhj9dj8" |
         participant == "4srlrzfe7zrkce" | participant == "h21fw4bmslr0vb" |
           participant == "7txj1h7m1zulc0" | participant == "tkp19u9btbvhee")

# since q id was not recorded on j for exp2 for some reason, we can only arbitrarily delete one of the two entries (according to the decision rule)
df_q_unique <- df_q %>%
  filter(!(jspsych_id == "a48zc72k379mm5" & StartDate == "2022-03-13 11:44:21" |
             jspsych_id == "hecbjunkhj9dj8" & StartDate == "2022-03-14 11:09:00" |
             jspsych_id == "4srlrzfe7zrkce" & StartDate == "2022-03-28 16:31:30" |
             jspsych_id == "h21fw4bmslr0vb" & StartDate == "2022-03-29 15:21:11" |
             jspsych_id == "7txj1h7m1zulc0" & StartDate == "2022-03-30 15:49:11" |
             jspsych_id == "tkp19u9btbvhee" & StartDate == "2022-04-07 19:30:53" |
             jspsych_id == "ept8xz3drgq38w" & StartDate == "2022-04-15 14:55:23" |
             jspsych_id == "w3uh3gtrykkjgc" & participant == "9143215998")) 
```

### Drop NA

This code chunk exists only because there were problems with the linking of Qualtrics and jspsych data. This code chunck is to filter out participants who did not do anything in the survey/categorization task. Some participants had two rows, one with Qualtrics response but blank jspsych response, the other with full jspsych response but blank Qualtrics response. This step is to merge those participants' Qualtrics and jspsych data into the same row.

```{r clean-na}
# see a list of q participants
df_q_id <- df_q_unique %>%
  select(jspsych_id, participant) 

# check if there are any j subs without q for exp2
df_j2_id %>%
  filter(!participant %in% df_q_id$jspsych_id)
# nope, luckily everyone on j had a q response in exp2

# yes q no j
na_j <- df_q_id %>%
  filter(is.na(jspsych_id) | jspsych_id == "22-0159")
# 22-0159 is someone who didn't pass the test and therefore wasn't able to 
# proceed to jspsych main task due to error (they were supposed to be able to)

# q-j mismatch (yes j; blank q survey for the row associated w the q id recorded on j)
na_q <- df_j1_id %>%
  filter(qualtrics_id %in% na_j$participant)
# there's a matching error bw q & j - in q, q id is there, but survey blank; the corresponding j id in q has another q id for the filled survey row
# there were 4 subs like this in exp1

# check for blank q rows with q id recorded in j 
na_j %>%
#  filter(participant %in% na_q$qualtrics_id) the following line of code does the same thing as this
  semi_join(na_q, by = c("participant" = "qualtrics_id"))

# check for q rows w j id that has response in j but blank q rows if matched by q id
df_q_id %>%
  filter(jspsych_id %in% na_q$participant) # turns out these responses were recorded under the same j id but different q id

# change their q id to the one recorded on j
df_q_unique <- df_q_unique %>%
  mutate(participant = case_when(jspsych_id == "wczbffevcxqk45" ~ "1044403354",
                                 jspsych_id == "en7lefm4g8mp17" ~ "4358939887",
                                 jspsych_id == "ghbk58xwj3w7v2" ~ "1475084897",
                                 TRUE ~ participant)) %>%
  filter(!is.na(jspsych_id)) # delete blank rows
# this person w jspsych_id == "5281kj47nhoboj", qualtrics_id == "6783315289" does not have a corresponding qualtrics survey filled out
# IDK Y! IDK WHO THAT IS! - how should we report this in the paper?

# for the rest of the blank q rows, I can't figure out which q id matches which j id, bc no q id was recorded on j in exp2
# so i just left them as they are
# doesn't matter bc we can join dfs using j id so we can ignore q

# manually code number of instruments they played
df_q_unique$Inst_num <- c(2, 1, 2, 3, 2, 1, 0, 2, 3, 2, 3, 0, 0, 3, 1, 2, 1, 0, 1, 1, 2, 1, 1, 1, 3, 1, 2, 2, 1, 3, 1, 1, 0, 2, 1, 1, 2, 0, 2, 0, 1, 1, 2, 2, 1, 2, 1, 7, 1, 2, 0, 0, 2, 1, 3, 1, 2, 2, 4, 0, 3, 0, 3, 3, 2, 2, 1, 0, 0, 1, 3, 4, 2, 2, 0, 3, 2, 4, 2, 1, 3, 2, 1, 2, 0, 3, 3, 1, 2, 2, 0, 2, 4, 1, 2, 2, 5, 1, 1, 1, 2, 2, 2, 0, 1, 1, 4)

# move column
df_q_unique <- df_q_unique %>%
  relocate(Inst_num, .after = Inst_yr)
```

### Practice and headphone test scores

The following code chunk extracts headphone test and practice test (training for major/minor categorization) scores and attaches them to the qualtrics demographic survey.

```{r practice-headphone}
# df for practice score
practice1 <- df_j1 %>%
  filter(designation == "PRACTICE-PASSED-SUMMARY") %>%
  select(participant, passed_practice, block_passed_practice, practice_score)

# df for headphone test
headphone1 <- df_j1 %>%
  filter(designation == "headphone-test") %>%
  select(participant, correct) %>%
  group_by(participant) %>%
  summarize(headphone = sum(correct)) 

# combine headphone test & practice test scores
practice_headphone1 <- left_join(headphone1, practice1)

# now do the same for exp2
practice2 <- df_j2 %>%
  filter(designation == "PRACTICE-PASSED-SUMMARY") %>%
  select(participant, passed_practice, block_passed_practice, practice_score)

headphone2 <- df_j2 %>%
  filter(designation == "headphone-test") %>%
  select(participant, correct) %>%
  group_by(participant) %>%
  summarize(headphone = sum(correct)) 

practice_headphone2 <- left_join(headphone2, practice2)

# combine headphone & practice test scores w demographics
df_q_plus_scores <- left_join(df_q_unique, rbind(practice_headphone1, practice_headphone2), by = c("jspsych_id" = "participant")) %>%
  mutate(passed_practice = factor(passed_practice, levels = c(0, 1), labels = c("No", "Yes")),
         block_passed_practice = factor(block_passed_practice),
         accuracy = practice_score/12)

# save a copy of clean q data
write_csv(df_q_plus_scores, "qualtrics_cleaned.csv")
```

## Cleaned data!

A snapshot of the data (aka codebook):

```{r codebook, echo = FALSE}
variable_names <- colnames(df_q_plus_scores)
variable_meaning = c("Time of participation", "Time when the survey response was recorded",
                     "Random 14-digit combination of letters and numbers, used to verify completion of jspsych task", "", "", 
                     "Year in university", 
                     "If not a student or faculty", "", 
                     "If not a psych or music major", 
                     "If they have ever played an instrument or not", 
                     "At what age they started playing an instrument", 
                     "Still playing now?", "List of instruments played", 
                     "Number of years they've played their most trained instrument", 
                     "Number of instruments played (including voice)", 
                     "List instruments used in exp1 that they've played (if any)",
                     "If they have ever participated in an ensemble or not", 
                     "Taken music courses or not", "List of music courses taken", 
                     "Can read music or not", 
                     "Has perfect pitch or not (1 = yes, 0 = no, -1 = don't know", 
                     "Time spent making music per week", 
                     "Time spent listening to music per day", 
                     "Number of concerts attended per year", 
                     "Proportion of each genre listened to (total 100)", 
                     "", "", "", "", "", "", "", "", "", "", 
                     "Other genres not mentioned",
                     "Random 10-digit Qualtircs ID",
                     "Self-rated ability to perceive and remember pitch", 
                     "Self-rated ability to perceive and remember tempo",
                     "headphone test score (out of 6)",
                     "Passed tonality categorization practice test or not", 
                     "How many attempts it took to pass the practice test (2 at most)",
                     "Practice test score (out of 12)",
                     "Number of correct answers in headphone test")
codebook <- data.frame(variable_names, variable_meaning)
kable(codebook)
```