---
title: "Eva Analyses"
author: "Eva Wu"
date: '`r Sys.Date()`'
output: github_document
---

## Statistical Analyses

Descriptives, exploratory plot

```{r descr}
# save data to be used
#attach(all)

# descriptives
#summary(cat_rtg)
#plot(instrument, pct_maj)
```

Correlations b/w 1) instrument's presumed valence, 2) mean explicit rating, and 3) tuning step & percent of major categorization, 
and 4)instrument's presumed valence & mean explicit rating

```{r cor}
#cor(cat_rtg$inst_id, cat_rtg$pct_maj)
#cor(cat_rtg$mean_rtg, cat_rtg$pct_maj)
#cor(cat_rtg$tuning_step, cat$pct_maj)
#cor(cat_rtg$inst_id, cat_rtg$mean_rtg)
```

Logistic regression

1) Percent major ~ instrument & tuning step

```{r log-reg}
#glm.fit <- glm(pct_maj ~ instrument + tuning_step, family = binomial) 
# family= binomial tells r to run logistic regression
#summary(glm.fit)
```

2) Percent major ~ mean explicit rating of each instrument & tuning step

```{r log-reg2}
#glm.fit2 <- glm(pct_maj ~ mean_rtg + tuning_step, family = binomial) 
# family= binomial tells r to run logistic regression
#summary(glm.fit2)
```

Linear regression

```{r lm-fit-both}
#lm.fit2 <- lm(pct_maj ~ tuning_step, data = cat)
#lm.fit <- lm(pct_maj ~ instrument + tuning_step, data = cat) 
#summary(lm.fit)
```

ANOVA exploring 1) whether adding instrument as a predictor significantly improves model, 
and 2) whether adding both predictors is significantly better than null model

```{r lm-anova}
#anova(lm.fit, lm.fit2) # adding instrument significantly improves model
#anova(lm.fit) # adding both predictors significantly better than null model
```
