---
title: "Instrument and Chord Perception"
author: "Stephen Van Hedger"
date: "22/02/2022"
output: github_document
---

```{r setup, include = FALSE}
library(lme4)
library(tidyverse)
library(effects)
library(ordinal)
library(emmeans)
library(lmerTest)
library(reshape2) # to pivot wider
library(corrplot)
library(rstatix)

knitr::opts_chunk$set(message = FALSE)
```

### Overview

Project examining how instrumental timbre influences categorization of three-note arpeggios as major versus minor. Middle note of the arpeggios has five steps, ranging from +0c (minor) to +100c (major) in 25c increments. Participants heard five instruments in total (randomized across trial): piano, oboe, violin, trumpet, xylophone.

Judgments should be theoretically independent of instrument. However, if certain timbres are more strongly associated with positive/negative affect, then it is possible that this will be reflected in the categorization task (i.e., "happier" instruments leading to a greater likelihood of responding "major"), despite no explicit instructions to base judgments on affective characteristics.

### 1. Load Data

```{r load, warning = FALSE}
data <- read_csv("inst-cat-uc-1.csv")
```

### 2. Generalized Linear Models

Here we report our main analyses

```{r model}
data.cat <- data %>% filter(designation == "MAIN-JUDGMENT") # extract cat data

#first, let's confirm that the inclusion of linear/quadratic/cubic effects is warranted via nested models
main.model3 <- lmer(selected_major ~ poly(tuning_step, 3)*instrument + (1+instrument|participant), data = data.cat)
main.model2 <- lmer(selected_major ~ poly(tuning_step, 2)*instrument + (1+instrument|participant), data = data.cat)
main.model1 <- lmer(selected_major ~ poly(tuning_step, 1)*instrument + (1+instrument|participant), data = data.cat)

anova(main.model3, main.model2) # strong evidence to keep cubic fit (versus just quadratic + linear)
anova(main.model3, main.model1) # strong evidence to keep cubic fit (versus just linear)
anova(main.model2, main.model1) # Eva: strong evidence to keep quadratic fit (versus just linear) - I guess that's a given when cubic is sig?

# cat results from the selected cubic model
summary(main.model3)

cat.emm <- emmeans(main.model3, "instrument")
pairs(cat.emm) # oboe < all others except piano; xylophone > piano

# assess explicit ratings of instruments
data.exp <- data %>% filter(designation == "INST-VALENCE-RTG")
data.exp$explicit_rtg <- ordered(data.exp$explicit_rtg)

explicit.model <- clmm(explicit_rtg ~ instrument + (1|participant), data = data.exp)
summary(explicit.model) # clear differences in reported capacity for instruments to play happy/sad
explicit.emm <- emmeans(explicit.model, "instrument")
pairs(explicit.emm) # violin < all others except oboe; xylophone > oboe
```

```{r summarize-pivot}
# how do categorization results map onto explicit ratings? 
# [aggregate = summarize (by mean) in dplyr]
cat.sum <- aggregate(selected_major ~ instrument*participant, FUN = mean, data = data.cat)
exp.sum <- aggregate(as.numeric(explicit_rtg) ~ instrument*participant, FUN = mean, data = data.exp)

#convert to wide format for correlational analyses [= pivot wider]
cat.sum.wide <- dcast(cat.sum, participant ~ instrument) 
exp.sum.wide <- dcast(exp.sum, participant ~ instrument)
cor.data <- cbind(cat.sum.wide[, c(2:6)], exp.sum.wide[, c(2:6)])
```

### 3. Cat-rtg correlation plot 

```{r corr-plot}
names(cor.data) = c("Oboe C", "Piano C", "Trumpet C", "Violin C", "Xylophone C", "Oboe E", "Piano E", "Trumpet E", "Violin E", "Xylophone E")
corvalues <- cor(cor.data) #cor matrix

col <- colorRampPalette(c("#4477AA",  "#77AADD", "#FFFFFF","#BB4444", "#EE9988"))
corrplot(corvalues, method = "color", col=col(200),    
         type = "upper", 
         addCoef.col = "black", # Add coefficient of correlation
         tl.col = "black", tl.srt = 45, # Text label color and rotation
         # Combine with significance
         # hide correlation coefficient on the principal diagonal
         diag = FALSE)
```

### 4. Categorization smooth plot

```{r smooth-plot}
eff.inst <- as.data.frame(effect(term = "poly(tuning_step,3)*instrument", mod = main.model3)) 
str(eff.inst) # shows structure of df
levels(eff.inst$instrument) <- c("Oboe", "Piano", "Trumpet", "Violin", "Xylophone", NA)

efplot.main <- ggplot(eff.inst, aes(x = tuning_step, y = fit, color = instrument)) +
  geom_point(size = 2) + 
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se), width = 0.1, lwd = 1) +
  #geom_line(size=1.2) +
  geom_smooth(formula = "y ~ poly(x, 3)", method = "glm", family = binomial, se = FALSE) +
  scale_x_continuous(labels = c("0", "25", "50", "75", "100")) +
  labs(y = "'Major' Response (Prop.)", x = "Tuning Step", color = "Instrument") +
  theme(plot.title = element_text(lineheight = .8, face = "bold", size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16),
        legend.box = "vertical", 
        legend.background = element_rect(color = NA),
        legend.position = c(0.20, 0.75),
        panel.background = element_rect(fill = "white", colour = "black"),
        panel.border = element_rect(color = "black", fill=NA, size=1.25),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        axis.text = element_text(size = 14),
        axis.line.x = element_line(colour = "black", size = 0.5, linetype = "solid"),
        axis.line.y = element_line(colour = "black", size = 0.5, linetype = "solid"),
        strip.text = element_text(size = 14))

png("inst-cat-plot.png", height = 4, width = 5.5, res = 300, units = "in")
efplot.main
dev.off()
```
